---
title: "Activity Tracker Data Analysis"
author: "Sachin B."
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Loading and preprocessing the data

### 1.1. Code for reading in the dataset 
```{r}

# extract "activity.csv" from "activity.zip" file
unzip("activity.zip",overwrite=TRUE)

# read data from "activity.csv" into "df" dataframe
df = read.csv("activity.csv")

# understanding data stored in df
str(df)

```

### 2.2. Code for processing the data ( Converting date from "Character" format to "Date" Format)
```{r}

df$date = as.Date(df$date,"%Y-%m-%d")

str(df)
```

## 2. What is mean total number of steps taken per day?

### 2.1. Remove incomplete observation
```{r}

df1=df[complete.cases(df), ]
str(df1)
```

### 2.2. Calculate the total number of steps taken per day
```{r}

daysteps = aggregate(steps ~ date, data = df1, sum)

head(daysteps)
    
```

### 2.3 Histogram of the total number of steps taken each day

```{r}

hist(daysteps$steps, xlab="Number of Steps", main="Histogram of Steps Per Day, Oct-Nov 2012")
    
```

### 2.4 The mean and median of the total number of steps taken per day

```{r}

daysteps_mean = mean(daysteps$steps)
cat("The mean of the total number of steps taken per day is :",daysteps_mean)

daysteps_median = median(daysteps$steps)
cat("\nThe median of the total number of steps taken per day is: ",daysteps_median)

```

<hr>

## 3. What is the average daily activity pattern?

### 3.1 Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}

avgpat = aggregate(steps ~ interval, data = df1, mean)
head(avgpat)

plot(avgpat$interval,avgpat$steps, type = "l" , xlab = "5-minute interval", ylab = "avg. no. of steps taken", main = "Average daily activity pattern")

```

### 3.2 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps

```{r}

# Learning Summary of 'avgpat' dataset
summary(avgpat)

# 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
avgpat[avgpat$steps == max(avgpat$steps),1]

```

## 4. Imputing missing values

### 4.1. Total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
# Total no. of rows
nrow(df)

# Total no. of rows without NA vales
    ## df1 = df[complete.cases(df), ]
nrow(df1)

# Total no. of rows with NAs

    ## na_records = nrow(df[!complete.cases(df), ])
 na_records = nrow(df) - nrow(df1)
 na_records

```


### 4.2. the mean for that 5-minute interval is filling in all of the missing values in the dataset

```{r}


# Identifying Positions of missing value in original dataset
na_pos <- which(is.na(df))
head(na_pos)


# Creating a new dataset that is equal to the original dataset but Replacing missing values with the mean for that 5-minute interval.

df2 <- df

for (i in na_pos) {
  index <- match(df[i,"interval"],avgpat$interval)
  df2[i,"steps"] <- avgpat[index,"steps"]
}

```

### 4.3 Histogram of the total number of steps taken each day
```{r}

df2steps = aggregate(steps ~ date, data = df2, sum)
head(df2steps)

hist(df2steps$steps, col="steelblue", xlab="Number of Steps", ylab="Number of Days",  main="Histogram of Steps Per Day, Oct-Nov 2012")

```

### 4.4 The mean and median of the total number of steps taken per day

```{r}

df2steps_mean = mean(df2steps$steps)
cat("The mean of the total number of steps taken per day is :",df2steps_mean)

df2steps_median = median(df2steps$steps)
cat("\nThe median of the total number of steps taken per day is: ",df2steps_median)

```

### 4.5  Values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}

# NA removed dataset
summary(daysteps$steps)

# dataset with NA imputed
summary(df2steps$steps)

```

In Activity dataset, steps per day mean and median doesn't affect much by replacing Na values with mean for that 5-minute slot. But 1st quartile and 3rd quartile values has been raise by around 1000 step counts which is helpful for having significant population for calculation.


## 5. Checking for differences in activity patterns between weekdays and weekends

### 5.1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
df2$week <- ifelse(weekdays(df2$date) %in% c("Saturday","Sunday"), "weekend", "weekday")

df2$week <- as.factor(df2$week)
                              
str(df2)

```

### 5.2. Panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}

weekavg <- aggregate(steps ~ interval + week, df2, mean )

library(ggplot2)

g <- ggplot(weekavg,aes(interval,steps, col = interval))

g+geom_line()+facet_grid(week~.) + labs(x = "Interval", y = "Number of steps", title = "Time Series Plot")

```